Inefficient Concessions and Mediation

### 2023-03-13 draft_narrow.tex
Update draft_narrow.tex
More minor tweaks to the intro.
China-US case added to intro
The China-US case is now worked into the intro with export controls on dual-use technology as a 3rd archetypal case -- along with disarmament and strategic territory.

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. This is true in non-violent disputes as well -- parties engage in costly conflict while they negotiate. If settlement can be achieved, additional costs of conflict -- including additional death and destruction -- are averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that conflict participants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

We introduce a two-sided signaling model in which concessions can provide future value to the recipient and/or serve as a signal that provides the recipient with information about the type of player the conceding state is. Using a mechanism design framework, we demonstrate that a mediator can facilitate more efficient concessions by removing uncertainty about the ability of the parties to commit to a settlement. Perhaps more importantly, the mediator can also sometimes facilitate settlement when the negotiating parties cannot achieve it through bilateral interaction.

Two archetypal situations illustrate the significance and difficulty of efficient concession-making in dispute settlement.  The first is the case of disarmament and the second is the concession of control over strategically valuable territory. In latter sections of the paper we discuss the applicability of the model to cases of nonviolent conflict as well, but peacemaking provides a particularly challenging and impactful context to illustrate the utility of inefficient concessions.

@@

Walter (1997) observes empirically that, due to the difficulty of making efficient disarmament concessions, civil wars tend to end on the battlefield, rather than through negotiated settlement. Interstate wars over territory are similarly prone to violent, rather than negotiated, resolution (Hensel 1996). Can mediation offer a path to settlement in these challenging contexts?

This paper revises our understanding of the reasons why efficient concessions are sometimes impossible to make, why inefficient concessions can sometimes be beneficial, and how mediators can both help increase the efficiency of concessions and make peace. In the war context, weapons or strategically valuable territory will not be conceded if the conceding side fears that territory or weapons may be used in future attacks against them. More broadly, countries are reluctant to make concessions that can be used against them in latter rounds of conflict or negotiation. One way to address this concern of dangerous concessions is to destroy the future value of the concession---e.g. decommission weapons or attempt to reduce the strategic value of the territory being conceded.  Another option, and the one on which we focus in this paper, is third-party mediation. We use this model to demonstrate that a third-party mediator can facilitate negotiated settlement by increasing trust and enabling disputants to make more efficient concessions, even when there is a risk those concessions might increase the bargaining power of their opponent in the future.

@@

Often, inefficient gifts occur in the formative stages of relationships or in immature relationships (Camerer 1988). Gifts given in established relationships, such as wedding presents or spousal gifts, are more likely to be cash gifts or efficient gifts that are specifically requested or selected from a registry. Since inefficient gift giving is more likely involved in nascent relationships and partnerships, it is natural to suppose that inefficient gifts play a signaling role in addition to the material value of transferring the gift's inherent worth.

>>>>

In the course of a military conflict, settlement negotiations often occur simultaneously with continued fighting. This is true in non-violent disputes as well, such as long-running episodes of geo-strategic competition -- parties engage in costly conflict while they negotiate. If settlement can be achieved, additional costs of conflict -- including additional death and destruction -- are averted. Third-party mediators often participate in such negotiations, but the literature remains divided regarding how and under what conditions mediation can be effective (e.g., Bercovitch 2000; Greig 2005a; Beardsley et al. 2006; Duursma 2020).  This paper sheds new light on the productive role mediators can play by examining their effect on the efficiency of the concessions that conflict participants can offer. In our context, the efficiency of a concession is determined by how much of the material value of the concession is intact when the recipient receives it. If the party who gives the concession destroys any of its material value, the concession is deemed to be less than fully efficient.

We introduce a two-sided signaling model in which concessions can provide future value to the recipient and/or serve as a signal that provides the recipient with information about the type of player the conceding state is. Using a mechanism design framework, we demonstrate that a mediator can facilitate more efficient concessions by removing uncertainty about the ability of the parties to commit to a settlement. Perhaps more importantly, the mediator can also sometimes facilitate settlement when the negotiating parties cannot achieve it through bilateral interaction. 
In other words, commitment is essentially a far-sighted preferences that allows cooperation to emerge, uncertainty can result in missed opportunities for cooperation, and concessions and mediation can facilitate conflict settlement.
%The sentence above literally parrots the editor's words back to them. Too on the nose?

%Attempting to add in the China-US case in a prominent way. The prose is more awkward than I would like still. - March 13, 2023 BG
Three archetypal situations illustrate the significance and difficulty of efficient concession-making in dispute settlement.  The first is the case of disarmament, the second is the concession of control over strategically valuable territory, and the third is the export of advanced technology. %In each case, efficient concessions are difficult to make because the item conceded may be used by the recipient to bolster their military capabilities, altering the balance of power in the ongoing conflict. Also true in each case, if dispute settlement is achieved, a public good is produced -- peace in the first two cases and innovation and growth in the civilian economy in the third case.

@@

In our third archetypal example, the US has recently moved to restrict the export of cutting-edge technology to China. China and the US are engaged in geo-strategic competition. Both countries could benefit from the more rapid innovation and growth that free trade in chips and other technology would enable, but there is risk that the technology exported will be used in the military sector to pursue advanced weapons, altering the military balance of power. As of 2023, trade in advanced technologies is restricted, reducing the global pace of innovation and economic growth.

Walter (1997) observes empirically that, due to the difficulty of making efficient disarmament concessions, civil wars tend to end on the battlefield, rather than through negotiated settlement. Interstate wars over territory are similarly prone to violent, rather than negotiated, resolution (Hensel 1996), and the trajectory of competition and cooperation between the US and China remains highly uncertain. Can mediation offer a path to settlement -- and a path to peace and prosperity -- in these challenging contexts?

This paper revises our understanding of the reasons why efficient concessions are sometimes impossible to make, why inefficient concessions can sometimes be beneficial, and how mediators can both help increase the efficiency of concessions and make peace. In the war context, weapons or strategically valuable territory will not be conceded if the conceding side fears that territory or weapons may be used in future attacks against them. More broadly, countries are reluctant to make concessions that can be used against them in latter rounds of conflict or negotiation. One way to address this concern of dangerous concessions is to destroy the future value of the concession---e.g. decommission weapons, attempt to reduce the strategic value of the territory being conceded, or only allowing trade in technologies without military applications. Another option, and the one on which we focus in this paper, is third-party mediation. We use this model to demonstrate that a third-party mediator can facilitate negotiated settlement by increasing trust and enabling disputants to make more efficient concessions, even when there is a risk those concessions might increase the bargaining power of their opponent in the future.

@@

%I need to add a bit about historical international relations, and a couple of cites, to the two paragraphs below. BG 3-13-2023
Often, inefficient gifts occur in the formative stages of relationships or in immature relationships (Camerer 1988). Gifts given in established relationships, such as wedding presents or spousal gifts, are more likely to be cash gifts or efficient gifts that are specifically requested or selected from a registry. Since inefficient gift giving is more likely involved in nascent relationships and partnerships, it is natural to suppose that inefficient gifts play a signaling role in addition to the material value of transferring the gift's inherent worth.
